Improving \textsl{Gaia} parallaxes with a data-driven model of the color--magnitude diagram

### 2017-03-14 paperDistance/paper.tex
2mass, refs, etc

\newcommand{\bibtex}{\textsc{Bib}\!\TeX} % bibtex. Not quite the correct typesetting, but close enough
\newcommand{\foreign}[1]{\textsl{#1}}
\newcommand{\acronym}[1]{{\small{#1}}}
\newcommand{\project}[1]{\textsl{#1}}
\newcommand{\tgas}{\project{\acronym{TGAS}}}
\newcommand{\tmass}{\project{\acronym{2MASS}}}
\newcommand{\gaia}{\project{Gaia}}
\newcommand{\xd}{\acronym{XD}}
\newcommand{\cmd}{\acronym{CMD}}

@@

Here we build and use just such a data-driven model.
In particular, we use all of the \gaia\ \tgas\ data, matched to the \tmass\
photometry, to make a model of the noise-free (or low-noise)
color--magnitude diagram (\cmd) of stars.
We use
an Empirical-Bayes approach known as Extreme Deconvolution (DWH CITE).
This method deconvolves heteroskedastic data to derive an estimate
of the noise-free or low-noise distribution that \emph{would have}
been observed with far better data.
This method has been used in astrophysics
previously to model the velocity distribution of stars in the Solar
Neighborhood (DWH CITE BOTH) and to perform quasar target selection
in \project{Sloan Digital Sky Survey} data (DWH CITE BOTH).
Its advantage over other deconvolution methods is that it takes as input,
and handles in a principled way, heteroskedastic data.
Its principle disadvantage for the present purposes is that it requires
or implies a strictly Gaussian noise model.
In the below, this requires us to transform the \cmd\ to a space in which
the noise is approximately Gaussian.

The novel use of \xd\ \citep{bovy11} here is to treat its output---the deconvolved \cmd---as
a prior for use in inference of individual-star distances.
These inferences, one per star, provide much more precise parallax, distance,
or absolute magnitude estimates than we get from each star's primary
\tgas\ parallax measurement alone.
That is, we are using the \xd\ model to de-noise the \gaia\ parallax
measurements.
Technically, since \xd\ produces a maximum-marginalized-likelihood estimator
of the deconvolved distribution,
its use as a prior is only approximately valid; it constitutes using the
data twice.
However, since the data set is large, the \xd\ approximation is not bad; it is
sometimes known as the ``empirical Bayes'' method, and is well studied (DWH REFS).

@@

\acknowledgments It is a pleasure to thank
  Jo Bovy (Toronto),
  Andy Casey (Monash),
and the attendees at the Stars Group Meeting at the Flatiron Institute
Center for Computational Astrophysics for comments and input.

This project was developed in part at the 2016 \acronym{NYC} Gaia Sprint, hosted
by the Center for Computational Astrophysics at the Simons Foundation
in New York City.

@@

[IS THERE A \tmass\ ACKNOWLEDGEMENT?]

>>>>

\newcommand{\bibtex}{\textsc{Bib}\!\TeX} % bibtex. Not quite the correct typesetting, but close enough
\newcommand{\foreign}[1]{\textsl{#1}}
\newcommand{\acronym}[1]{{\small{#1}}}
\newcommand{\project}[1]{\textsl{#1}}
\newcommand{\tgas}{\project{\acronym{TGAS}}}
\newcommand{\tmass}{\project{\acronym{2MASS}}}
\newcommand{\gaia}{\project{Gaia}}
\newcommand{\panstarrs}{\project{Pan\acronym{STARRS}}}
\newcommand{\xd}{\acronym{XD}}
\newcommand{\cmd}{\acronym{CMD}}

@@

Here we build and use just such a data-driven model.  In particular,
we use all of the \gaia\ \tgas\ data, matched to the
\tmass\ photometry (\citealt{skrutskie06}), and in the footprint of
the \panstarrs\ Dust Map (CITE GREEN), to make a model of the
noise-free (or low-noise) color--magnitude diagram (\cmd) of stars.
We use an Empirical-Bayes approach known as Extreme Deconvolution (XD;
\citealt{bovy11}).
This method deconvolves heteroskedastic data to derive an
estimate of the noise-free or low-noise distribution that \emph{would
  have} been observed with far better data.
This method has been used in astrophysics
previously to model the velocity distribution of stars in the Solar
Neighborhood (DWH CITE BOTH) and to perform quasar target selection
in \project{Sloan Digital Sky Survey} data (DWH CITE BOTH).
Its advantage over other deconvolution methods is that it takes as input,
and handles in a principled way, heteroskedastic data.
Its principle disadvantage for the present purposes is that it requires
or implies a strictly Gaussian noise model.
In the below, this requires us to transform the \cmd\ to a space in which
the noise is approximately Gaussian.

The novel use of \xd\ here is to treat its output---the deconvolved \cmd---as
a prior for use in inference of individual-star distances.
These inferences, one per star, provide much more precise parallax, distance,
or absolute magnitude estimates than we get from each star's primary
\tgas\ parallax measurement alone.
That is, we are using the \xd\ model to de-noise the \gaia\ parallax
measurements.
Technically, since \xd\ produces a maximum-marginalized-likelihood estimator
of the deconvolved distribution,
its use as a prior is only approximately valid; it constitutes using the
data twice.
However, since the data set is large, the \xd\ approximation is not bad; it is
sometimes known as the ``empirical Bayes'' method, and is well studied (DWH REFS).

@@

\acknowledgments It is a pleasure to thank
  Ana Bonaca (Harvard),
  Andy Casey (Monash),
and the attendees at the Stars Group Meeting at the Flatiron Institute
Center for Computational Astrophysics for comments and input.
This project was developed in part at the 2016 \acronym{NYC} Gaia Sprint, hosted
by the Center for Computational Astrophysics at the Simons Foundation
in New York City.

@@

This publication makes use of data products from the Two Micron All
Sky Survey, which is a joint project of the
University of Massachusetts and the Infrared Processing and Analysis
Center/California Institute of Technology, funded by the National
Aeronautics and Space Administration and the National Science
Foundation.
